---
title: "Habitat-species associations"
subtitle: "Given a number of tree species and habitat types in a plot, are species significantly aggregated within habitats?"
author: "Mauro Lepore"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Habitat-species associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(digits = 3)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  out.width = "70%",
  fig.align = "center",
  fig.height = 9,
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```

This article shows how to determine habitat-species associations with the function [`tt_test()`](https://forestgeo.github.io/fgeo.analyze/reference/tt_test), developed by Sabrina Russo, Daniel Zuleta, Matteo Detto, and Kyle Harms. 

## Setup

First, install and load ("open") the relevant packages.

```R
# install.packages("remotes")
remotes::install_github("forestgeo/fgeo")
```

```{r}
library(dplyr)
library(ggplot2)
library(fgeo)
library(fgeo.data)
```

```{r}
# For reproducible results
set.seed(1014)
```

## Load census and habitat data

We will use example datasets that come with __fgeo.analyze__.

```{r}
census <- luquillo_tree6_random

str(census)
```

```{r}
# Creating habitat data from elevation data
habitat <- fgeo_habitat(luquillo_elevation, gridsize = 20, n = 4)
str(habitat)
```

```{r}
fgeo.plot::autoplot(habitat)
```


To load your own data, you may run something like this:

```R
load("PATH/CENSUS_DATA.rdata")
census <- CENSUS_DATA

load("PATH/HABITAT-DATA.rdata")
habitat <- HABITAT-DATA
```

## Pick data

We will pick alive trees, of 10 mm or more, and of sufficiently abundant species.

```{r}
pick <- filter(
  census,
  # Keep only alive
  status == "A", 
  # Keep dbh of 10 mm or more (drops missing dbh)
  dbh >= 10
)
# Count number of rows per species
pick <- add_count(pick, sp)
# Keep sufficiently abundant trees
pick <- filter(pick, n > 50)

# Summary
unique(select(pick, sp, n))
```

## Overview

Before testing, we can overview the relationship between species and habitats with a plot.

```{r}
# Tweaks
offset <- 20 / 2
habitat2 <- mutate(
  habitat, 
  # Center species and habitat data
  x = gx + offset, 
  y = gy + offset,
  # From continuous to categorical
  habitats = as.factor(habitats)
)
```

```{r, out.width="90%", fig.widh=(6 * 0.9 / 0.7)}
ggplot(pick, aes(x = gx, y = gy)) +
  geom_raster(data = habitat2, aes(x, y, fill = habitats)) + 
  geom_point() +
  coord_fixed() +
  facet_wrap(~sp) +
  labs(fill = "Habitat")
```

## `tt_test()` and any number of species

```{r}
tt_test_result <- tt_test(pick, habitat)
tt_test_result
```

To help you interpret the results, you can use `summary()`.

```{r}
summary(tt_test_result)
```

You may want to combine the output into a single matrix by row-binding each element of the results-list.

```{r}
Reduce(rbind, tt_test_result)
```

You also can gather all results into a single dataframe -- this lets you use a wide range of tools for data manipulation and visualization.

```{r}
to_df(tt_test_result)
```

You can benefit from storing your results in a dataframe. Compared to a matrix, a dataframe fits better in common workflows for data manipulation and visualization. The dataframe is the most important data structure used in __dplyr__, __ggplot2__, and many other packages. Here are some examples of what you can do with our dataframe output. (The next few code chunks use [the pipe operator (`%>%`)](http://style.tidyverse.org/pipes.html) to avoid saving intermediary results and to make our code more expressive -- where each line is an imperative statement that communicates our intention.)

* Pick interesting rows and sort them in a meaningful way.

```{r}
to_df(tt_test_result) %>% 
  filter(sp == "CASARB")
```

* Summarize

```{r}
to_df(tt_test_result) %>% 
  group_by(sp, habitat) %>% 
  summarize(total_stem_count = sum(N.Hab))
```
